Mathematical Language Processing \\ Project

### 2014-06-30 paper.tex
typo & some quirky sentences rewritten

Mathematical formulae are viable sources of information for a wide range of
scientists. Often, they contain identifiers whose meaning might be at first
unknown or at least ambiguous to the reader (depending on their knowledge).
Therefore, one usually needs to study the surrounding text to find the
relevant definition. An automatic information retrieval system can be used to
reduce the reader's effort by displaying the most relevant definition relation
found, to the reader. Students and scientists of other disciplines would
especially profit from a system that helps them to understand formulae more
quickly. In the long term, the extracted identifier definition tuples contribute
to an increased machine readability of scientific publications. This builds a
foundation for added value services such as search, clustering and improved
accessibility.

To build such a system, a labeled text corpus that annotates identifiers and
their definition is desirable. At the project start, such a corpus was not
available.
Consequently we had to start manual investigation of individual articles.
Our first observation was many identifier definitions use a fixed string pattern
to explain the definition to the reader. Furhtermore most
definitions are usually very close to the occurence of the identifier.
Thus we calculate the probablities for correct identifier definition tuples based
on part-of-speech (POS) tagged based distance metrics.
This correlates to the experience that readers usually extract identifier definitions from context
that is given by the surrounding text. In combination with the Stratosphere
system \cite{Alexandrov2010} that is capable of scalable data processing, we
rapidly generate a large amount of definition relation candidates with only
minimal implementation overhead for the parallelization.

As the target text corpus we chose the Wikipedia because of two facts. First,
most articles make use of \texttt{<math/>} tags (\emph{texvc} as an input
language) for formulae. The identification of these tags are easy to find,
and from the MathML output, it is easy to extract the identifiers.
Second, the text is already annotated with mark-up.
Particularly, hyperlinks to other articles within Wikipedia are of
interest as they typically wrap around any number of words and indicate that
these in combination are relevant in the given context or (respectively) sentence.

The English Wikipedia contains roughly four million articles. Even if we only
pick articles containing \texttt{<math/>} tags, our processor still needs to
compute with tens of thousands of articles. Especially when using text
annotators (e.g., \emph{POS} tagger \cite{Rathna96}), like Stanford's NLP
framework, one can make use of a parallel processing system to speed up
computation. Therefore we implement the proposed strategy with the
Stratosphere system based on the PACT programming model \cite{Alexandrov2010}.

@@

At first, we implemented a simple identifier definition extractor that is
based on a set of static patterns. As this is a fairly robust approach and easy
to implement, it serves as a good reference point in terms of
performance. It simply iterates through the text, trying to find word groups,
that are matched by a pattern. The patterns being used to discover description
terms are depicted in table \ref{tpatterns}. Due to the fact that we already
tokenized an annotated the articles in a previous step in the MLP system, we
can make use of POS tags here as well.

Note, determiners not only contain articles, but also quantifiers and
distributives. The last pattern in table \ref{tpatterns} contains `*/DT'. This
is a shorthand for every word, that has the POS tag `DT' (determiner), will be
matched at this position. Otherwise this pattern whould be rather large, as it
needs to contain every possible determiner. \emph{identifer} as well as
\emph{description} are placholders, that mark the positions of the entities
from a possible defintion relation.

@@

Extracting relevant identifiers from the article rely on the assumption that
the author will use \texttt{<math/>} tags for all formulae. This said, a
formula that is written within the text cannot be recognized, and therefore
can never be extracted by our system.

The fact that we can estimate all relevant identifiers for an article (see
Section \ref{ir}), combined with some common assumptions about the relation,
can be exploited to largely reduce the set of candidates that need to be
ranked. Please note that this reduction is essential for retrieving the
correct relations for our approach. Otherwise almost any word would be ranked
and the precision of the retrieval would drop significantly.

The basic assumption of our approach is that the two entities of a definition
relation co-occur in the same sentence. In other words, if we want to retrieve
the description for an identifier, in the set of sentences that contain the
identifier the description should be found. Thus said, any other sentences can
be ignored. Furthermore, we assume that it is more likely to find the
description in first sentences than in the latter. This is based on the idea
that authors introduce the meaning of an identifer and than subsequent use it,
without necessarily repeating the meaning.

Another assumption can be made about the lexical class of the description we
want to rank. They are nouns or even noun phrases (e.g., \texttt{`the
effective spring constant k'} or \texttt{`mass m of something'}). We discard
all other words (according to their POS tag) except noun phrases and Wikipedia
hyperlinks. These are the candidates for a definition relation. Noun phrases
and hyperlinks may consist of multiple words. For all intents and purposes, it
is not necessary to threat noun phrases and hyperlinks as a set of words, and
therefore, will be threated subsequently as if they were one. This is
important, due to the fact that the overall ranking will be greatly influenced
by the distance of candidates to the position of the identifer.

@@

Throughout our experiments, we made some observations that had an impact on the
accuracy of retrieving the correct set of identifiers. First of all, people
tend to incorrectly use \emph{texvc} when trying to create formulas. For example,
\texttt{\textbackslash text\{log\}} is more often used than the correct
operator \texttt{\textbackslash log}. Another problem is that sometimes people
use indices as a form of `in field' annotation, like $T_{before}$ and
$T_{after}$. The identifier $T$ is defined in the surrounding text, but
neither $T_{before}$ nor $T_{after}$. There are more ambiguities. For example
the superscripted 2 in $x^{2}$ and $\sigma^{2}$ can be interpreted as the
power or as a part of the identifier. Another ambiguity is that the
multiplication sign can be omitted, so that it is undecidable for a naive
program whether $ab^{2}$ contains one or two identifiers.

We took a very conservative approach and preprocessed all formulas.
The \TeX~command \texttt{\textbackslash text\{\}} blocks along with subscriptions containing
more than a single character will be removed before analysis. Superscripts
will also be ignored in terms of being a part of the identifier. Moreover, we
created a comprehensive blacklist to improve the results further. Short words
like `a', `A', and `I', which are very common in the English language, could be
easily matched by our processor in the surrounding text if a formula contains
one of those identifiers, and therefore, will also be blacklisted. Additionally,
we blacklist common mathematical operators, constants, and functions.

Because we do not have any annotated test corpora, evaluation has to be
performed by hand. Therefore, we took a sample of 30 random articles and
counted all matches. The resulting estimates for the identifier retrieval
performance are: \emph{Recall: 0.99} and \emph{Precision: 0.86}, which satisfy
our information needs, as we are mostly interested in recall at this stage.

@@

There is no gold standard evaluation datasets to measure the performance of
identifier definition extractors. Thus we created on on our own dataset. This
is a very time consuming job. At the moment, the dataset only contains two
large articles (revision ids are included) with around 100 identifier
definitions. This dataset is also available on the project repository.

@@

Our experiments showed that selecting candidates according to their POS tags
combined with numerical statistics about the text surface, can lead to quality
results. However, this approach is only applicable under certain conditions.
For identifiers which are seldom seen, our statistical approach tends to fail.
In that situation, other methods, especially supervised ones, are preferred.
Unfortunately, many of them require a labeled test corpus to measure the
performance of a classifier that could be trained with our generated data.
Currently, we are planning to use the NTCIR-Math-10 Task, Math Understanding
Subtask gold standard dataset \cite{overview} for a comparable evaluation.

>>>>

Mathematical formulae are viable sources of information for a wide range of
scientists. Often, they contain identifiers whose meaning might be at first
unknown or at least ambiguous to the reader (depending on their knowledge).
Therefore, one usually needs to study the surrounding text to find the
relevant definition. An automatic information retrieval system can be used to
reduce the reader's effort by displaying the most relevant definition relation
found to the reader. Students and scientists of other disciplines would
especially profit from a system that helps them to understand formulae more
quickly. In the long term, the extracted identifier definition tuples contribute
to an increased machine readability of scientific publications. This builds a
foundation for added value services such as search, clustering and improved
accessibility.

To build such a system, a labelled text corpus that annotates identifiers and
their definition is desirable. At the project start, such a corpus was not
available. Consequently we had to start manual investigation of individual
articles. Our first observation was that many identifier definitions use a
fixed string pattern to explain the definition to the reader. Furthermore,
most definitions usually appear very close to the related identifier in the
sentences. Thus, we calculate the probabilities for correct identifier
definition tuples based on distance metrics for certain part-of-speech (POS)
tagged words. This correlates to the experience that readers usually extract
identifier definitions from context that is given by the surrounding text.

We chose the Wikipedia as the target text corpus because of two facts. First,
most articles make use of \texttt{<math/>} tags (\emph{texvc} as an input
language) for formulae. The identification of \texttt{<math/>} tags is
trivial, and from the MathML output, it is easy to extract the identifiers.
Second, the articles are already annotated with mark-up. Particularly, hyperlinks
to other articles within Wikipedia are of interest as they typically wrap
around any number of words and indicate that these in combination are relevant
in the given context or (respectively) sentence.

The English Wikipedia contains roughly four million articles. Even if we only
pick articles containing \texttt{<math/>} tags, our processor still needs to
compute with tens of thousands of articles. Especially when using text
annotators (e.g., \emph{POS} tagger \cite{Rathna96}), like Stanford's NLP
framework, one can make use of a parallel processing system to speed up
computation. We implement the proposed strategy with the Stratosphere system
\cite{stratosphere}. It is based on the PACT programming model
\cite{Alexandrov2010}, which enables us to rapidly generate a large amount of
definition relation candidates with only minimal implementation overhead for
the parallelization.

@@

At first, we implemented a simple identifier definition extractor that is
based on a set of static patterns. As this is a fairly robust approach and easy
to implement, it serves as a good reference point in terms of
performance. It simply iterates through the text, trying to find word groups,
that are matched by a pattern. The patterns being used to discover description
terms are depicted in Table \ref{tpatterns}. Due to the fact that we already
tokenized and annotated the articles in a previous step in the MLP system, we
can make use of POS tags here as well.

Note, determiners not only contain articles, but also quantifiers and
distributives. The last pattern in Table \ref{tpatterns} contains `*/DT'. This
is a shorthand for every word, that has the POS tag `DT' (determiner).
Otherwise this pattern would be rather large, as it needs to contain every
possible determiner. \emph{IDENTIFIER} as well as \emph{DESCRIPTION} are
place-holder, that mark the positions of the entities from a possible definition
relation.

@@

Extracting relevant identifiers from the article relies on the assumption that
the author will use \texttt{<math/>} tags for all formulae. This said, a
formula that is written in the running text cannot be recognized, and therefore,
cannot be extracted by our system.

The fact that we can estimate all relevant identifiers for an article (see
Section \ref{ir}), combined with some common assumptions about definition relations,
can be exploited to largely reduce the set of candidates that need to be
ranked. Please note that this reduction is essential for retrieving the
correct relations for our approach. Otherwise almost any word would be ranked
and the precision of the retrieval would drop significantly.

The basic assumption of our approach is that the two entities of a definition
relation co-occur in the same sentence. In other words, if we want to retrieve
the description for an identifier, only sentences containing the identifier
could include the definition relation. Having said this, any other sentences can
be ignored. Furthermore, we assume that it is more likely to find the
description in first sentences than in the latter. This is based on the idea
that authors introduce the meaning of an identifier and than subsequent use the
identifier, without necessarily repeating its definition.

Another assumption can be made about the lexical class of the definition relation we
want to rank. The descriptions are nouns or even noun phrases (e.g., \texttt{`the
effective spring constant k'} or \texttt{`mass m of something'}). We discard
all other words (according to their POS tag) except noun phrases and Wikipedia
hyper-links. These are the candidates descriptions for a definition relation. Noun phrases
and hyper-links may consist of multiple words. For all intents and purposes, it
is not necessary to threat noun phrases and hyper-links as a set of words, and
therefore, they will be treated subsequently as if they were one. This is
important, due to the fact that the overall ranking will be greatly influenced
by the distance of candidates to the position of the identifier.

@@

Throughout our experiments, we made some observations that had an impact on the
accuracy of retrieving the correct set of identifiers. First of all, people
tend to use \emph{texvc} only as a typesetting language and neglect its semantic capabilities. For example,
\texttt{\textbackslash text\{log\}} is more often used than the correct
operator \texttt{\textbackslash log}. Another problem is that sometimes people
use indices as a form of `in field' annotation, like $T_{before}$ and
$T_{after}$. The identifier $T$ is defined in the surrounding text, but
neither $T_{before}$ nor $T_{after}$. There are more ambiguities. For example
the superscripted 2 in $x^{2}$ and $\sigma^{2}$ can be interpreted as the
power or as a part of the identifier. Another ambiguity is that the
multiplication sign can be omitted, so that it is undecidable for a naive
program whether $ab^{2}$ contains one or two identifiers.

We took a very conservative approach and preprocessed all formulas. The
\TeX~command \texttt{\textbackslash text\{\}} blocks along with subscriptions
containing more than a single character will be removed before analysis.
Superscripts will also be ignored in terms of being a part of the identifier.
Moreover, we created a comprehensive blacklist to improve the results further.
Identifier like `a', `A', and `I', which are also very common in the English
language, could be easily matched by our processor in the surrounding text,
and therefore, will also be blacklisted. Additionally, we blacklist common
mathematical operators, constants, and functions.

We took a sample of 30 random articles and counted all matches by hand. The
resulting estimates for the identifier retrieval performance are
\emph{Recall: 0.99} and \emph{Precision: 0.86}, which satisfy our information
needs, as we are mostly interested in recall at this stage.

@@

At the start of our project there were no gold standard datasets available to
measure the performance of identifier definition extractors. Thus, we created
one on our own. This is a very time consuming job. At the moment, the dataset
only contains two large articles (revision ids are included) with around 100
identifier definitions. This dataset is also available on the project
repository.

@@

Our experiments showed that selecting candidates according to their POS tags
combined with numerical statistics about the text surface, can lead to quality
results. However, this approach is only applicable under certain conditions.
For identifiers which are seldom seen, our statistical approach tends to fail.
In that situation, other methods, especially supervised ones, are preferred.
Unfortunately, many of them require a labelled test corpus to measure the
performance of a classifier that could be trained with our generated data.
Currently, we are planning to use the NTCIR-Math-10 Task, Math Understanding
Subtask gold standard dataset \cite{overview} for a comparable evaluation.
